Solving the maximum independant set problem with Tropical tensor networks

### 2021-05-25 paper/paper.tex
update paper

\section{Computing degeneracy}
\section{Utilizing the sparsity}
Tensor network compression is an important tool to utilize sparsity.

@@

\subsection{The equivalence between branching and compression}
We are going the verify the Lemmas used in branching algorithm in book~\cite{Fomin2013}.
\begin{theorem} % basic
  If a vertex $v$ is in an independent set $I$, then none of its neighbors can be in $I$.
On the other hand, if $I$ is a maximum (and thus maximal) independent set,
and thus if $v$ is not in $I$ then at least one of its neighbors is in $I$.
\end{theorem}
\begin{theorem} % 2.6
Let $G=(V,E)$ be a graph, let $v$ and $w$ be adjacent vertices of $G$ such that $N[v] \subseteq N[w]$. Then
\begin{equation}
\alpha(G)=\alpha(G\backslash w).
\end{equation}
\end{theorem}

\begin{theorem} % 2.7
  Let $G = (V, E)$ be a graph and let $v$ be a vertex of $G$. If no maximum
independent set of $G$ contains $v$ then every maximum independent set of $G$ contains
at least two vertices of $N(v)$.
\end{theorem}

\begin{theorem} % 2.8
Let $G = (V, E)$ be a graph and $v$ a vertex of $G$. Then
\begin{equation}
\alpha(G) = \max(1 + \alpha(G \backslash N[v]), \alpha(G \backslash (M(v) \cup \{v\})).
\end{equation}
\end{theorem}

>>>>

\section{Computing degeneracy}
Consider the following 5 node graph.

\begin{figure}[H]
    \centerline{\includegraphics[width=0.4\columnwidth,trim={0 0cm 0 0},clip]{../notebooks/fig1.pdf}}
\end{figure}

If put a vector of size $2$ at each vertex
\begin{equation}
    \left(\begin{matrix}
        1 \\
        x
    \end{matrix}\right),
\end{equation}

and a matrix of size $2 \times 2$ at each edge
\begin{equation}
    \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right).
\end{equation}

\begin{figure}[H]
    \centerline{\includegraphics[width=0.4\columnwidth,trim={0 0cm 0 0},clip]{../notebooks/fig2.pdf}}
\end{figure}

Here $x$ is a variable. Then we contract this graph with einsum \texttt{a,b,c,d,ab,ad,bc,bd,cd,ce->}.
The result is a scalar function of $x$, it corresponds to the independance polynomial~\cite{}.
\begin{equation}
I(G, x) = \sum_{k=1}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the the degeneracy for independant set size $k$.

\subsection{Symbolic computing}
The most straight forward approach is the symbolic computing, where we store a polynomial of $x$ as a vector of factors.
We define the algebra between these vectors of factors and insert them to the original tensor network contraction algorithm.
However, this method suffers from a space overhead that propotional to the maximum independant set size.

\subsection{Polynomial fitting}
Let $m$ be the maximum independent set size and $X$ be a set of $m+1$ random real numbers, e.g. $\{0, 1, 2, \ldots, m\}$.
We compute the einsum contraction for each $x \in X$ and obtain the following relations
\begin{align}
a_0 + a_1 x_1 + a_1 x_1^2 + \ldots + a_m x_1^m &= y_0\\
a_0 + a_1 x_2 + a_2 x_2^2 + \ldots + a_m x_2^m &= y_1\\
\ldots&\\
a_0 + a_1 x_m + a_2 x_m^2 + \ldots + a_m x_m^m& = y_m
\end{align}
The polynomial fitting between $X$ and $Y = \{y_0, y_1, \ldots, y_m\}$ gives us the factors.

\subsection{Fourier transformation}
\begin{align}
\left(\begin{matrix}
1 & x_1 & x_1^2 & \ldots & x_1^m \\
1 & x_2 & x_2^2 & \ldots & x_2^m \\
\vdots & \vdots & \vdots &\ddots & \vdots \\
1 & x_m & x_m^2 & \ldots & x_m^m
\end{matrix}\right)
\left(\begin{matrix}
a_0 \\ a_1 \\ \vdots \\ a_m
\end{matrix}\right)
= \left(\begin{matrix}
y_0 \\ y_1 \\ \vdots \\ y_m
\end{matrix}\right)
\end{align}

\begin{align}
\left(\begin{matrix}
1 & r\omega & r^2\omega^2 & \ldots & r^m\omega^m \\
1 & r\omega^2 & r^2\omega^4 & \ldots & r^m\omega^{2m} \\
\vdots & \vdots & \vdots &\ddots & \vdots \\
1 & r\omega^m & r^2\omega^{2m} & \ldots & r^m\omega^{m^2}
\end{matrix}\right)
\left(\begin{matrix}
a_0 \\ a_1 \\ \vdots \\ a_m
\end{matrix}\right)
= \left(\begin{matrix}
y_0 \\ y_1 \\ \vdots \\ y_m
\end{matrix}\right)
\end{align}

When $r=1$, the left side is a DFT matrix. We can obtain the factors using the relation $\vec a = {\rm FFT^{-1}}(\omega) \cdot \vec y$.
In the special case that $\omega = e^{-2\pi i/(m+1)}$, it is directly solvable with inverse fast fourier transformation algorithm in package \texttt{FFTW}.

\begin{equation}
{\rm FFT}(\omega) \cdot \vec a_r = \vec y
\end{equation}
where $(\vec a_r)_k = a_k r ^k$, by choosing diferent $r$, we can obtain better precision in low independant set size region  ($\omega<1$) and high independant set size region ($\omega>1$).

\subsection{Finite field algebra}

\section{Utilizing the sparsity}
Tensor network compression is an important tool to utilize sparsity.

@@

\subsection{The tensor network compression detects branching rules automatically}

\textit{max-size rule} The resulting tensor of Tropical tensor contraction stores only the local independant set size that maximizes the internal degrees on freedom.
Where the internal degree of freedom are edges that does not connect to tensors not contracted.

\textit{max-interior rule} The resulting tensor of Tropical tensor contraction stores only the local independant set size that maximizes the internal degrees on freedom.

We are going the verify the Lemmas used in branching algorithm in book~\cite{Fomin2013}.
\begin{theorem}\label{rule:one} % basic
  If a vertex $v$ is in an independent set $I$, then none of its neighbors can be in $I$.
On the other hand, if $I$ is a maximum (and thus maximal) independent set,
and thus if $v$ is not in $I$ then at least one of its neighbors is in $I$.
\end{theorem}

Contract $N[v]$, each entry of the resulting tensor of rank $|N(v)|$ corresponds to a locally maximized independant set size with fixed boundary configuration.
If the boundary configuration is a bit string of 0s, $\sigma_v$ will takes value $1$ to maximize the local independant set size.

\begin{theorem} % 2.5
Let $G=(V,E)$ be a graph, let $v$ and $w$ be adjacent vertices of $G$ such that $N[v] \subseteq N[w]$. Then
\begin{equation}
\alpha(G)=\alpha(G\backslash w).
\end{equation}
\end{theorem}

Contract $N[w]$, both $\{v, w\}$ disapear from the tensor indices because they are inner degrees of freedom.
If $w$ is one, then $N[v]$ are all zeros, the resulting tensor element can not be larger than setting $v=1$ and $w=0$.
By the maximization rule, the local tensor does not change if we remove $w$.

\begin{theorem} % 2.6
  Let $G = (V, E)$ be a graph and let $v$ be a vertex of $G$. If no maximum
independent set of $G$ contains $v$ then every maximum independent set of $G$ contains
at least two vertices of $N(v)$.
\end{theorem}

\begin{theorem} % 2.7
Let $G = (V, E)$ be a graph and $v$ a vertex of $G$. Then
\begin{equation}
\alpha(G) = \max(1 + \alpha(G \backslash N[v]), \alpha(G \backslash (M(v) \cup \{v\})).
\end{equation}
\end{theorem}

Here, $M(v)$ is the set of mirrors of $v$ in $G$.
A vertex $w \in N^2(v)$ is called a mirror of $v$ if $N(v) \backslash N(w)$ is a clique.
This rule states that if $v$ is not in $M$, there exists an MIS $I$ that $M(v)\notin I$.
otherwise, there must be one of $N(v)$ in the MIS (maximization rule).
If $w$ is in $I$, then none of $N(v) \cap N(w)$ is in $I$, then there must be one of node in the clique $N(v)\backslash N(w)$ in $I$ (maximization rule),
since clique has at most one node in the MIS, the tensor compression will eliminate this solution by moving the occuppied node to the interior.
Hence, the tensor compression rule do captures the mirror rule.
